Neural Power Units

### 2020-06-02 papers/main.tex
clean up; get tikz

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
\usepackage[preprint]{neurips_2020}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

@@

% figures
\graphicspath{{./../plots/}}
\usepackage{tikz}
\usetikzlibrary{arrows,backgrounds,positioning,intersections}
\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepgfplotslibrary{groupplots}
\usepgfplotslibrary{polar}
\usepgfplotslibrary{smithchart}
\usepgfplotslibrary{statistics}
\usepgfplotslibrary{dateplot}

\newcommand{\real}{\text{Re}}
\newcommand{\imag}{\text{Im}}
\newcommand{\wre}{w_{r}}
\newcommand{\wim}{w_{i}}
\newcommand{\Wre}{\bm W_{r}}
\newcommand{\Wim}{\bm W_{i}}

@@

\title{Neural Power Unit \\-- A Bayesian Approach To Neural Arithmetic}
\author{
  Niklas Heim,
  V\'aclav \v Sm\'idl,
  Tom\'a\v s Pevn\'y \\
  Artificial Intelligence Center\\
  Czech Technical University\\
  Prague, CZ 120 00\\
  \texttt{\{niklas.heim, vasek.smidl, tomas.pevny\}@aic.fel.cvut.cz}\\
}

@@

\begin{abstract}
  Common Neural Networks can approximate simple arithmetic operations, but fail
  to generalize beyond the range of numbers that were seen during training. A
  new class of units called \emph{Neural Arithmetic Units} aim to overcome this
  difficulty, but are limited either to operate on positive numbers (NALU by
  \citet{trask_neural_2018}), or can only represent simple addition and
  multiplication (NAU \& NMU by \citet{madsen_neural_2020}).  We introduce a
  Neural Arithmetic Unit that operates on the full domain of real numbers and
  is capable of learning arbitrary power functions called \emph{Neural Power
  Unit} (NPU). The NPU enables not only the learning simple arithmetic
  operations, but is also capable of learning e.g. Taylor expansions with very
  few parameters, which makes the NPU highly explainable.
  \footnote{Implementation of neural arithmetic units:
  \url{https://github.com/nmheim/NeuralArithmetic.jl}}.
\end{abstract}

@@

Numbers and simple algebra are essential not only to human intelligence but
also to the survival of many other
species~\citep{dehaene_number_2011,gallistel_finding_2018}.  This can be taken
as a hint that arithmetic is an important ingredient to a successful,
intelligent agent.  State of the art neural networks are capable of learning
simple arithmetic, but they fail to extrapolate beyond the ranges seen during
training~\citep{suzgun_evaluating_2018,lake_generalization_2018}.  The
inability to generalize to unseen inputs is a fundamental problem that hints at
a lack of \emph{understanding} of the given task. The model merely memorizes
the seen inputs and fails to abstract the true learning task.  The failure of
numerical extrapolation on simple arithmetic tasks has been shown
by~\cite{trask_neural_2018}, who also introduce a new class of \emph{Neural
Arithmetic Units} that show good extrapolation performance on some arithmetic
tasks.

The \emph{Neural Arithmetic Logic Unit} (NALU) can learn addition,
subtraction, multiplication, and division with two severe limitations: the
inputs that the NALU is trained with have to be positive, and should be
significantly larger than zero, otherwise the layer will not converge to
a correct solution.

\cite{madsen_neural_2020} introduce two new arithmetic layers, the \emph{Neural
Addition Unit} (NAU) and the \emph{Neural Multiplication Unit} (NMU).
Stacked, they can learn addition, subtraction, and multiplication and they converge
for any input range. However, the NMU cannot learn division.

Including Neural Arithmetic Units in common neural networks, apart from the
potential increase in extrapolation capabilities, has additional advantages.
They promise to reduce the amount of parameters that are needed for a given
task, can drastically improve the explainability of our models (more on the
explainability of the NPU in Sec.~\ref{sec:neural_power_unit}).  Additionally,
the field of Scientific Machine Learning, which, in part, is concerned with
learning differential equations~\citep{rackauckas_universal_2020} could benefit
from the inductive bias neural arithmetic.

\subsection*{Our Contribution}%
\label{sub:our_contribution}

We introduce a new arithmetic layer called the \emph{Neural Power Unit} (NPU)
which is capable of learning addition, subtraction, and arbitrary power
functions, which includes multiplication and division.
Our approach is inspired by the NALU, but overcomes its limitations to
non-small positive numbers by leveraging the complex plane.

@@

\subsection{Neural Arithmetic Logic Unit}%
\label{sub:neural_arithmetic_logic_unit}

\citet{trask_neural_2018} have demonstrated how severe the extrapolation
problem is even for the simplest arithmetic operations, such as summing or
multiplying two numbers.  In order to increase the power of abstraction of NNs
they propose a \emph{Neural Arithmetic Logic Unit} (NALU) which is capable of
learning arithmetic addition, subtraction, multiplication, and division
$\{+,-,\times,\div\}$ with stunning extrapolation accuracy.  However, the NALU
comes with the severe limitation not being able to handle negative inputs due
to the logarithm in the multiplication part of the NALU:

\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x
                          & \bm W& = \tanh(\hat{\bm W}) \odot \sigma(\hat{\bm M}) \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm W(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}
Additionally the NALU has severe convergence problems for small inputs due to
a non-smooth loss surface close to zero, as shown by \cite{madsen_neural_2020}.

\todo{should this be here already?}
With the simple exemplary task of learning a function $f:\mathbb R^2 \rightarrow \mathbb R^2$
\begin{equation}
  \label{eq:xx_xdivy}
  f(\bm x) = f(x,y) = [xy,x/y] = \bm t
\end{equation}
we can demonstrate the strengths and weaknesses of the NALU.  The NALU only
learns successfully if it is trained on positive numbers, so we train with
samples from $\bm x \in \mathcal U^2(0.1,2)$ and plot the error $\epsilon =
|\bm t - \hat{\bm t}|^2$ in Fig.~\ref{fig:simple_err}.

\subsection{Neural Multiplication Unit}%
\label{sub:neural_multiplication_unit}

A multiplication layer that can handle small and negative inputs was introduced
by \citet{madsen_neural_2020}.  The \emph{Neural Multiplication Unit} (NMU) is
defined by Eq.~\ref{eq:nmu} and is typically used in conjunction with the
so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i M_{ij} z_{i} + 1 - M_{ij}  &M_{ij}=\min(\max(M_{ij}, 0), 1)\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm A \bm x &A_{ij}=\min(\max(A_{ij}, 0), 1)
\end{align}
In both NMU and NAU the weights are clipped to $[0,1]$, and typically regularized
with $\mathcal{R}$:
\begin{align}
  \label{eq:rsparse}
  \mathcal{R} = \sum_{ij} \min(W_{ij}, 1-W_{ij})
\end{align}

The combination of NAU and NMU can thus learn $\{+,-,\times\}$, but no division.
This is shown in Fig.~\ref{fig:simple_err} as well. The NMU
can perfectly represent the multiplication of the two inputs, but fails at
learning division.

\todo{put this in acknowledgements
all the results in this paper were create with the help of the following Julia
packages: \cite{rackauckas_differentialequationsjl_2017} + (Flux.jl)}

\begin{figure}
  \centering
  \includegraphics[width=0.245\linewidth]{simple_err_add.pdf}
  \includegraphics[width=0.245\linewidth]{simple_err_mult.pdf}
  \includegraphics[width=0.245\linewidth]{simple_err_div.pdf}
  \includegraphics[width=0.245\linewidth]{simple_err_sqrt.pdf}
  \caption{Comparison of different models learning the function
  $(x,y)\rightarrow(xy,x/y)$ with $(x,y)\in\mathcal U^2(0.1,2)$.  Each
  plot shows the mean-squared error on a logarithmic color scale ($\hat{\bm
  t}=\text{model}(\bm x)$).  The Dense network learns both multiplication and
  division, but fails to extrapolate. NALU learns both tasks perfectly, but
  completely fails if one of the inputs is negative.  The NMU successfully
  learns multiplication, but not division. Only the NPU learns both tasks and
  can extrapolate with an error that is less than $0.01$ in most places.}%
  \label{fig:simple_err}
\end{figure}
\todo{add in depth experiment desc. of fig.~\ref{fig:simple_err} in appendix}
\begin{table}
  \centering
  \caption{Comparison of different $2\times2$ layers on the task
  $(x,y)\rightarrow(xy,\frac{x}{y})$.  $(x,y) \in \mathcal U^2(-2,2)$.}
  \label{tab:simple_err}
  \input{simple_err.tex}
\end{table}

@@

Eq.~\ref{eq:nalu_mult} enables the NALU to learn multiplication. If the weight
matrix $\bm W$ would not be restricted to values in the range $(-1,1)$ it could
in principle learn arbitrary power functions, because
\begin{align}
  z = \exp(w\log x) = x^w \text{ for } x>0.
\end{align}

However, this would complicate the convergence issues of the NALU even more.
We therefore suggest to leverage the power of the complex plane, by using the
complex logarithm and by promoting $w$ to a complex number.  This will enable
us not only to lift the positivity constraint on $x$, but also provides and
additional dimension that helps to avoid regions with an uninformative gradient
signal\todo{rephrase this. do we have an educating example where it is obvious
that complex plane helps? the stuff in fig.1 does not work as well with
non-complex NPU, but I have not idea how to plot that}.  Introducing a complex
weight matrix somewhere in a larger network would mean that the gradients in
other layers would potentially become complex as well, which would result in
converting the whole network to complex numbers.  This seems like a high price
to pay, but fortunately there is a way around this.  We can output only the
real part of the result $\real(z)$, but still maintain a complex $w$ by
treating it as a sum of real and imaginary part $w=\wre+i\wim$. We are thus
interested in
\begin{align}
  \label{eq:exp_w_log_complex}
  \real(z) &= \real(\exp(w\log x)),
\end{align}
with a real input $x$. The complex logarithm is defined as $\log z=\log r +
i\theta$, which, for real inputs $x$, can be simplified to $\log z = \log r +
ik\pi$, where $k\in \{0,1\}$. With that Eq.~\ref{eq:exp_w_log_complex} can be
rewritten to
\begin{align}
  \real(z) &= \real(\exp((\wre + i\wim) (\log r + ik\pi))) \\
    &= \exp(\wre\log r - \wim k\pi) \cos(\wim\log r + k\pi\wre),
\end{align}
where we used Euler's formula\footnote{Euler's formula states that for any real
number $x$: $re^{ix} = \cos x + i\sin x$}.
\todo{explicitly compute gradients?}

\subsection{Neural Power Unit (NPU)}%
\label{sub:neural_power_unit_npu_}

\begin{figure}
  \centering
  \resizebox{.3\textwidth}{!}{\input{../plots/exp_log_sin/log_comp.tikz.tex}}
  \resizebox{.3\textwidth}{!}{\input{../plots/exp_log_sin/exp_comp.tikz.tex}}
  \resizebox{.3\textwidth}{!}{\input{../plots/exp_log_sin/sin_comp.tikz.tex}}
  \caption{Function learning comparison of different layers.}%
  \label{fig:sin_comp}
\end{figure}

The \emph{Neural Power Unit} (NPU) with two weight matrices $\Wre$ and $\Wim$
then reads
\begin{align}
  \bm r = |\bm x|,
  \hspace{1cm}
  & k_i = \begin{cases}
     0  & x_i \leq 0 \\
    \pi & x_i > 0
  \end{cases}
\end{align}
\begin{align}
  \bm z &= \exp(\Wre \log\bm r - \Wim\bm k) \odot \cos(\Wim\log \bm r + \Wre\bm k)
\end{align}
where $\bm k$ is a vector that is zero where $\bm x$ is positive and $\pi$
where it is negative, $\bm r$ is the element-wise absolute value.
In the final network we are only interested in real powers, so we recommend to
strongly regularize the imaginary part of the weights $\Wim$.

\subsection{Gated NPU}%
\label{sub:gated_npu}
As pointed out by \cite{madsen_neural_2020}, the loss surface of the NALU, and
thus also the NPU, is not always smooth. For some input and weight combinations
it can happen that the loss close to zero becomes very large as shown in
Fig.~\ref{fig:task_loss_npu_nmu}.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_loss_npu_nmu.pdf}
  \caption{asdf}%
  \label{fig:task_loss_npu_nmu}
\end{figure}
However, most applications are dealing with batches of data which average out
these cases, as shown in Fig.~\ref{fig:task_loss_npu_nmu_average} for a batch
size of 32.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_loss_npu_nmu_average.pdf}
  \caption{asdfadsf}%
  \label{fig:task_loss_npu_nmu_average}
\end{figure}

In practice we encountered difficulties of obtaining sparse solutions with the
NPU. If the inputs to an NPU unit are small, the gradient becomes practically
zero, as we can see in Fig.~\ref{fig:id_loss_npu_nmu}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{id_loss_npu_nmu.pdf}
  \caption{learning identity with small inputs}%
  \label{fig:id_loss_npu_nmu}
\end{figure}

\section{Experimetns}%
\label{sec:experimetns}
This can be demonstrated by training a traditional, dense NN to approximate
three functions
\begin{align}
  \label{eq:approx_tasks}
  f(x) = e^x, && g(x) = \log(x), && h(x) = \sin(x)
\end{align}

Neural Arithmetic Units aim to overcome this problem with units that are able
to represent simple arithmetic operations exactly. By composition of these
units it becomes possible to learn e.g. exponentials, logarithms, and power
functions. This promises to improve on the extrapolation capabilities of NNs
even on tasks that are beyond artificial arithmetic tasks such as learning
multiplication, or a certain power function \todo{cite some examples}.

\subsection{10 param func}%
\label{sub:10_param_func}

>>>>

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
\usepackage{neurips_2020}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsthm}         % definitions
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{caption}
\usepackage{mathtools}

\newtheorem*{definition}{Definition}

@@

% tikz
\usepackage{tikz}
\usetikzlibrary{%
  arrows,%
  backgrounds,
  intersections,
  shapes,
  shapes.misc,% wg. rounded rectangle
  shapes.arrows,%
  chains,%
  matrix,%
  positioning,% wg. " of "
  scopes,%
  decorations.pathmorphing,% /pgf/decoration/random steps | erste Graphik
  shadows,%
  patterns% for hatched rect
}

% plots
\graphicspath{{./../plots/}}
\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepgfplotslibrary{groupplots}
\usepgfplotslibrary{polar}
\usepgfplotslibrary{smithchart}
\usepgfplotslibrary{statistics}
\usepgfplotslibrary{dateplot}

\newcommand{\real}{\text{Re}}
\newcommand{\imag}{\text{Im}}
\newcommand{\wre}{w_{r}}
\newcommand{\wim}{w_{i}}
\newcommand{\Wre}{\bm W_{r}}
\newcommand{\Wim}{\bm W_{i}}
\newcommand{\npu}{\text{NPU}}
\newcommand{\naivenpu}{\text{NaiveNPU}}
\newcommand{\nau}{\text{NAU}}
\newcommand{\nmu}{\text{NMU}}
\newcommand{\nalu}{\text{NALU}}
\newcommand{\inalu}{\text{iNALU}}
\newcommand{\dense}{\text{Dense}}
\newcommand{\chain}{\text{Chain}}

@@

\title{Neural Power Units}
\author{
  Niklas Heim,
  Tom\'a\v s Pevn\'y,
  V\'aclav \v Sm\'idl\\
  Artificial Intelligence Center\\
  Czech Technical University\\
  Prague, CZ 120 00\\
  \texttt{\{niklas.heim, vasek.smidl, tomas.pevny\}@aic.fel.cvut.cz}\\
}

@@

\begin{abstract}
  Conventional Neural Networks can approximate simple arithmetic operations, but fail
  to generalize beyond the range of numbers that were seen during training.
  \emph{Neural Arithmetic Units} aim to overcome this difficulty, but current
  arithmetic units are either limited to operate on positive numbers or can
  only represent a subset of arithmetic operations. We introduce the \emph{Neural
  Power Unit} (NPU) that operates on the
  \emph{full domain of real numbers} and is capable of \emph{learning
  arbitrary power functions} in a single layer.
  The NPU thus fixes the shortcomings of existing arithmetic units and extends
  their expressivity. This is achieved by using formulas from complex arithmetics
  without requiring a conversion of the network to complex numbers.
  We show where the NPU outperforms its
  competitors in terms of accuracy and sparsity on artificial arithmetic
  datasets. Additionally, we demonstrate how the NPU can be used as a highly
  interpretable model to discover the generating equation of a dynamical system
  purely from the data.
\end{abstract}

@@

Numbers and simple algebra are essential not only to human intelligence but
also to the survival of many other
species~\citep{dehaene_number_2011,gallistel_finding_2018}. A successful,
intelligent agent should therefore be able to perform simple arithmetic.  State
of the art neural networks are capable of learning arithmetic, but they
fail to extrapolate beyond the ranges seen during
training~\citep{suzgun_evaluating_2018,lake_generalization_2018}.  The
inability to generalize to unseen inputs is a fundamental problem that hints at
a lack of \emph{understanding} of the given task. The model merely memorizes
the seen inputs and fails to abstract the true learning task.  The failure of
numerical extrapolation on simple arithmetic tasks has been shown
by~\cite{trask_neural_2018}, who also introduce a new class of \emph{Neural
Arithmetic Units} that show good extrapolation performance on some arithmetic
tasks.

Including Neural Arithmetic Units in common neural networks, promises to
significantly increase their extrapolation capabilities due to their inductive
bias towards numerical computation. This is especially true for tasks in which
the data contains mathematical relationships.  Additionally, they promise to
reduce the amount of parameters that are needed for a given task, and can
therefore drastically improve the explainability of the model.  We demonstrate
how neural arithmetic units as part of a \emph{Neural Ordinary Differential
Equation} (NODE, \cite{chen_neural_2019}) can outperform a dense network with
significantly fewer parameters in Sec.~\ref{sub:fractional_sir_model}.  In
fact, our NPU can be used to directly read out the correct generating ODE from
the fitted model.  This is inline with recent efforts to build
\emph{interpretable} models instead of \emph{explaining} black box models
\citep{rudin_stop_2019}, like conventional neural networks.

The currently available arithmetic units all have different strengths and
weaknesses, but none of them solve simple arithmetic completely.
The \emph{Neural Arithmetic Logic Unit} (NALU) by \cite{trask_neural_2018},
chronologically, was the first arithmetic unit. It can solve
addition ($+$, including subtraction), multiplication ($\times$), and division
($\div$), but is limited to positive inputs. The convergence of the
NALU is quite fragile due to an internal gating mechanism between addition and
multiplication paths as well as the use of a logarithm which is problematic for
small inputs.  Recently, \cite{schlor_inalu_2020} introduced the \emph{improved NALU} (iNALU,
to fix the NALU's shortcomings. However, it significantly increases its complexity and we
observe only a slight improvement in performance.\\
\cite{madsen_neural_2020} solve ($+,\times$) with two new units: the
\emph{Neural Addition Unit} (NAU), and the \emph{Neural Multiplication Unit}
(NMU). Instead of gating between addition and multiplication paths, they are
separate units that can be stacked. They can work with the full range real
numbers, converge much more reliably, but are not capable of representing
division.

\subsection*{Our Contributions}%
\label{sub:our_contribution}

\textbf{Neural Power Unit.}
We introduce a new arithmetic layer (NPU,
Sec.~\ref{sec:neural_power_unit}) which is capable of learning power
functions $(\prod x_{i}^{w_i})$ for arbitrary real input $x_i$ and power $w_i$ thus including multiplication ($x_1\times x_2 = x_1^1 x_2^1$) as well as division $(x_1\div x_2=x_1^{1}x_2^{-1})$ . This is achieved by using formulas from complex arithmetics (Sec.~\ref{sub:naive_npu}).
Stacks of NAUs and NPUs can thus learn the full spectrum of simple arithmetic operations.

\textbf{Convergence improvement.}
We address the known convergence issues of neural arithmetic units by introducing a
\emph{relevance gate} that smooths out the loss surface of the NPU
(Sec.~\ref{sub:npu}). With the relevance gate the NPU reaches extrapolation and
sparsity errors that are on par with the NMU on ($\times$) and outperform NALU
on ($\div,\sqrt{\cdot}$).

\textbf{Interpretability.}
We show how the NPU can be used as a highly interpretable model for discovery of epidemiological models. 
Specifically, we demontrate its ability to provide a model that can be interpreted as fractional SIR model (Sec.~\ref{sub:fractional_sir_model})
that was used to fit the COVID-19 outbreak in various countries
\citep{taghvaei_fractional_2020}.

@@

A number of different approaches to automatically solve arithmetic tasks have
been studied in the recent years. Approaches include Neural GPUs
\citep{kaiser_neural_2016}, Grid LSTMs \citep{kalchbrenner_grid_2016}, Neural
Turing Machines \citep{graves_neural_2014}, and Neural Random Access Machines
\citep{kurach_neural_2016}. They solve tasks like binary addition and
multiplication, or single digit arithmetic. The Neural Status Register
\citep{faber_neural_2020} focusses on control flow.  The Neural Arithmetic
Expression Calculator \citep{chen_neural_2018}, a hierarchical reinforcement
learner is the only method that solves the division problem, but it operates on
character sequences of arithmetic expressions. Related is symbolic integration
with transformers \citep{lample_deep_2019}. Unfortunately, most of the named models
have severe problems with extrapolation \citep{madsen_measuring_2019,saxton_analysing_2019}.
A solution to the extrapolation problem could be Neural Arithmetic Units. They are
designed with an inductive bias towards systematic, arithmetic computation.
However, currently they are limited in their capabilities of expressing the full
range of simple arithmetic operations ($+,\times,\div$).
In the following two sections we briefly describe the currently available
arithmetic layers including their advantages and drawbacks.

\subsection{Neural Arithmetic Logic Units}%
\label{sub:neural_arithmetic_logic_unit}

\citet{trask_neural_2018} have demonstrated the severity of the extrapolation
problem of dense networks for even the simplest arithmetic operations, such as
summing or multiplying two numbers.  In order to increase the power of
abstraction to for arithmetic tasks they propose the \emph{Neural Arithmetic
Logic Unit} (NALU) which is capable of learning ($+,\times,\div$).  However,
the NALU cannot handle negative inputs correctly due to the logarithm in
Eq.~\ref{eq:nalu_mult}:
\begin{definition}[{\bf NALU}]
  The NALU consits of a ($+$) and a ($\times$) path that share their weights $\bm{\hat{W}}$ and $\bm{\hat{M}}$.
\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x
                          & \bm{\hat W}& = \tanh(\bm{W}) \odot \sigma(\bm{M}) \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm{\hat W}(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}
\end{definition}
Additionally, the logarithm destabilizes training to an extent that the chance of success
can drop below 20\% for ($+,\times$), it becomes practically impossible to
learn ($\div$), and difficult to learn from small inputs in general
\citep{madsen_measuring_2019}.
\cite{schlor_inalu_2020} provide a detailed description of the shortcomings of
the NALU and they suggest an \emph{improved NALU} (iNALU).  The iNALU addresses
the NALU's problems through a number of mechanisms.  It has independent
addition and multiplication weights for Eq.~\ref{eq:nalu_add} and
Eq.~\ref{eq:nalu_mult}, clips weights and gradients to improve training
stability, regularizes the weights to push them away from zero, and, most
importantly, introduces a mechanism to recover the sign that is lost due to the
absolute value in the logarithm. Additionally, the authors propose to
reinitialize the network if its loss is not improving during training.  We
include the iNALU in one of our experiments and find that it only slightly
improves the NALU's performance (Sec.~\ref{sub:simple_arithmetic_task}) at the
cost of a significantly more complicated unit. Our NPU avoids all these mechanisms
by internally using complex arithmetics.

\subsection{Neural Multiplication Unit \& Neural Addition Unit}%
\label{sub:neural_multiplication_unit}

Instead of trying to fix the NALU's convergence issues
\cite{madsen_neural_2020} propose a new unit for ($\times$) only.
The \emph{Neural Multiplication Unit} (NMU) uses explicit multiplications and learns
to gate between identity and ($\times$) of inputs. The NMU is
defined by Eq.~\ref{eq:nmu} and is typically used in conjunction with the
so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{definition}[{\bf NMU \& NAU}]
  NMU and NAU are two units that can be stacked to model ($+,\times$).
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i \hat M_{ij} z_{i} + 1 - \hat M_{ij}  & \hat M_{ij}=\min(\max(M_{ij}, 0), 1)\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm{\hat{A}} \bm x & \hat A_{ij}=\min(\max(A_{ij}, -1), 1)
\end{align}
\end{definition}
Both NMU and NAU are regularized with $\mathcal{R} = \sum_{ij} \min(|W_{ij}|,
|1-W_{ij}|)$ and their weights are clipped, which biases them towards learning
an operation or pruning it completely.  The combination of NAU and NMU can thus
learn $(+,\times)$ for both positive and negative inputs. Training NAU and NMU
is stable and succeeds much more frequently than with the NALU, but they cannot
represent ($\div$), which we address with our NPU.

@@

To fix the deficiencies of current arithmetic units, we propose a new
arithmetic unit (inspired by NALU) that can learn arbitrary power functions $(x^w)$ (including
$\times,\div$) for positive and negative numbers, and still train well.
Combined with the NAU we solve the full range of arithmetic operations.
This is possible through a simple modification of the ($\times$)-path of the
NALU (Eq.~\ref{eq:nalu_mult_2}).  We suggest to use the complex logarithm and
to allow $\bm W$ to become complex.  The complex logarithm is also defined for
negative inputs and a complex $\bm W$ aides convergence (see
Sec.~\ref{sub:fractional_sir_model}). The improvement during training might be
explained by the additional imaginary parameters that make it possible to avoid
regions with an uninformative gradient signal.

\subsection{Naive Neural Power Unit -- NaiveNPU}%
\label{sub:naive_npu}

With the modifications introduced above we can extend the multiplication path
of the NALU from
\begin{align}
  \label{eq:nalu_mult_2}
  \bm m = \exp \bm W(\log_{\text{real}}(|\bm x|+\epsilon))
\end{align}
to use the complex logarithm ($\log\coloneqq\log_{\text{complex}}$) and a
complex weight $\bm W$ to
\begin{align}
  \label{eq:1}
  \bm z = \exp(\bm W\log \bm x) = \exp\left((\bm \Wre + i\Wim) \log\bm x\right).
\end{align}
The complex log in Eq.~\ref{eq:1} lifts the positivity constraint on $\bm x$ resulting
in a layer that can process both positive and negative numbers correctly.  A complex
weight matrix $\bm W$ somewhere in a larger network would result in complex
gradients in other layers.  This would effectively result in doubling the
number of parameters of the whole network. As we are only interested in real
networks outputs, we can avoid this doubling by considering only the real part
of the output $\bm z$:
\begin{align}
  \real(\bm z) &= \real(\exp((\Wre + i\Wim)(\log\bm r + i\pi\bm k))) \\
    \label{eq:npumult}
    &= \exp(\Wre\log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log\bm r + \pi\Wre\bm k).
\end{align}
Above we have used Euler's formula\footnote{Euler's formula states that for any
real number $x$: $e^{ix} = \cos x + i\sin x$} and the fact that the complex
logarithm for real valued inputs is
\begin{align}
  \log x = \log r + i\theta = \log r + ik\pi,
\end{align}
where $k=0$ if $r\geq0$ and $k=1$ if $r<0$.
A diagram of the NaiveNPU is shown in Fig.~\ref{fig:npu_diagram}.
\begin{definition}[{\bf NaiveNPU}]
  The Naive Neural Power Unit with matrices $\Wre$ and
  $\Wim$ representing real and imaginary part of the complex numbers defined as
  \begin{gather}
    \label{eq:naivenpu_def}
    \bm z = \exp(\Wre \log\bm r - \Wim\bm k) \odot \cos(\Wim\log \bm r + \Wre\bm k), \text{ where }\\
    \nonumber
    \bm r = |\bm x|,
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
       \pi & x_i > 0
    \end{cases}
  \end{gather}
\end{definition}

\begin{figure}
  \begin{minipage}{.45\textwidth}
      \resizebox{0.9\textwidth}{!}{\input{../tikz/npu-diagram.tex}}
      \captionof{figure}{NaiveNPU diagram, with input $\bm x$ and output $\bm y$.
      Vectors in green, trainables in orange, functions in blue.}
      \label{fig:npu_diagram}
  \end{minipage}
  \hspace{0.01\textwidth}
  \begin{minipage}{0.54\textwidth}
    \resizebox{\textwidth}{!}{\input{../tikz/gatednpu-diagram.tex}}
    \captionof{figure}{NPU diagram. The NPU has a relevance gate $\bm g$
    (hatched background) in front of the input to the actual unit to prevent
    vanishing gradients.}
    \label{fig:gatednpu_diagram}
  \end{minipage}
\end{figure}

\subsection{The Relevance Gate -- NPU}%
\label{sub:npu}

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the identity on one of two inputs
(neglecting the second one):
\begin{align}
  \nonumber
  \mathcal{L} &= |\text{m}(x_1,x_2) - x_1|, \quad \text{where}
  & \text{m} = \npu(2,1), \quad
  &x_1 \sim \mathcal U(0,2), \quad x_2 \sim \mathcal U(0,0.05).
\end{align}
The left plot in Fig.~\ref{fig:id_loss} depicts
the the norm of gradient of a batch of two dimensional inputs, one of which is small and
irrelevant. It is clearly visible that even for this simple example, a large
part of the gradient surface of the NaiveNPU is zero.  This vanishing gradient can be explained
as follows. One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  We therefore introduce a gate on the input of the NPU that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.
\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm z = \exp(\Wre \log\bm r - \Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \Wre\bm k), \text{ where } \\
    \bm r = \bm{\hat g} \odot |\bm x| + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
      \hat g_i\pi & x_i > 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1)
  \end{gather}
\end{definition}
\begin{figure}
  \centering
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=\textwidth]{npu_gatednpu_id_grad.pdf}
  \end{minipage}
  \begin{minipage}{0.42\textwidth}
  \end{minipage}
  \caption{Absolute value of a gradient of NaiveNPU and NPU for the task
  of learning the identity on $x_1$. Inputs and
  loss are defined on the right, gradient surfaces on the left
  (black areas are beyond the color scale). The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  for central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they will adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same loss for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as L1.  In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as an interpretable model.

\subsection{Initialization}%
\label{sub:initialization}
We recommend to initialize the NPU with a Glorot Uniform distribution on the
real weights $\Wre$. The imaginary weights $\Wim$ can be initialized to zeros,
so they will only be used where necessary, and the gate $\bm g$ with 0.5, so
the NPU can choose to output 1.
\todo{should this be longer? or not a section at all?}

\section{Experiments}%
\label{sec:experiments}

In Sec.~\ref{sub:fractional_sir_model} we show how the NPU can be used as a
highly interpretable model which can extract the generating equation of an ODE
that contains fractional powers.
In Sec.~\ref{sub:simple_arithmetic_task}\,\&\,\ref{sub:large_scale_arithmetic_task}
we provide experiments on standard arithmetic tasks in which the NPU outperforms
the NALU in all tasks.

\subsection{Equation Discovery of an Epidemiological Model}%
\label{sub:fractional_sir_model}
\todo{other titles: Discovering the Equations of an Epidemic}
